\pkg{digiRhythm}: an \proglang{R} package for rhythmic behavior assessment in livestock

### 2023-05-15 jstatsoft/article.Rnw
amendements

\section[Introduction]{Introduction: The Degree of Functional Coupling in the context of livestock production} \label{sec:intro}

%new start
The real-time monitoring of animal behaviour is becoming increasingly important in agriculture. It can provide new opportunities to enhance the efficiency and sustainability of livestock production and is used to enhance animal welfare (Wemelsfelder et al., 2001). Animal behaviour can be altered by stress, pain or disease, but monitoring these changes in welfare is still a challenge, since some behavioural definitions are unclear and not specifically linked to a unique adverse event (Theurer et al., 2013; Levitis et al., 2009). Relevant information about animal welfare can be obtained by studying individual differences in the capacity for behavioural adaptation (Cooper et al., 2005). Indeed, during the animals life, significant changes can occur due to internal and external stimuli; these alterations arouse adaptive behavioural patterns that can be monitored by analysing the rhythmicity of behaviours (Berger et al., 2003).
Due to the increasing development of ICT (Information and Communication Technology) based on small, non-invasive sensors, large amounts of animal behaviour data are regularly recorded, and can be used to calculate parameters describing biological rhythms (Theurer et al., 2013; Scheibe and Gromann, 2006; Berger et al. 2003). Scheibe et al. (1999) developed a parameter able to quantify and investigate the rhythmicity of behavioural patterns, the Degree of Functional Coupling (DFC), which is based on telemetric measurements of activity and/or feeding. The DFC provides an estimate of synchronization between an individual and its environment. It is calculated to each 24 h period within rolling seven day periods and predicts values between 0%
and 100% (Scheibe et al., 1999, Berger et al., 2003). An organism with a high DFC is highly synchronized with its environment, while lower DFC's indicate a lower synchronization. Breakdowns in the DFC response can be caused by stressors (e.g. transport, disease, etc.) that affect the time structure of behaviour (Berger et al., 2003; Scheibe et al., 1999). In fact, animals that live in stress-free conditions show activities all tuned to one zeitgeber (i. e. 24 hour periodicity), while their activity rhythm will deviate from the zeitgeber frequency if they are subjected to stressful situations (Scheibe et al., 1999; Berger et al., 2003).
Previous studies demonstrated that calves change their behaviour as illness develops (reduced feeding, less social behaviour and increased rest) as an adaptive response to cope with the disease (Owen-Ashley et al., 2006, Dantzer and Kelley, 2007; Borderas et al. 2008). A better understanding of changes in behaviour due to illness or stress of handling is valuable to improve animal welfare and productivity as well as handler safety (Grandin et al., 1989; Kilgore et al., 2005; Borderas et al., 2008; Robért et al., 2011). Bio-rhythmic analysis has the potential to provide the basis for the early detection of psychological and physiological disorders (Berger et al., 1999).
%new end

suggest to delete start:
With the progression of digital technologies livestock production has changed significantly. Sensors are now regularly mounted on animals, with the intention of improving management practices through monitoring performance, such as daily milk yields. However, existing sensor systems are also used for the early detection of physiological changes. To date, the reliability of sensor based abnormality detection still proves potential for optimization. While processes on farms, including livestock production, are increasingly automated, it is still not possible to reliably assess the welfare state of a farm animals. The Degree of functional coupling (DFC) presents a unique approach to activity evaluation. The DFC is a metric used to describe the extent to which a measured activity is significantly cyclic to 24 hours and thus harmonically synchronized with the periodicity of the environment. Such activity can be, e.g., the number of steps during a time interval or an acceleration measured with an accelerometer mounted on the animal, but could also be the visiting times of feeding troughs or watering stations. It is assumed that organisms in a good welfare state have a rhythmicity close to 1, whereas the rhythmicity of the organism drops when its physiological state is compromised. The DFC was first introduced by \cite{sinz1976systemanalyse} to gain insight into the circadian rhythm of sheep. Hereby, the authors confirmed that the DFC of sheep dropped before they died.  \cite{scheibe1999comparative} further used the  DFC to evaluate the welfare of zoo animals and wildlife by applying a sequence of mathematical operations on continuous signals collected via animal-borne sensors.
suggest to delete end:

@@

The current package has been developed for farm animals, but the principles of rhythmicity apply for any organism, including humans. Therefore we highly recommend future research to evaluate the DigiRhythm package in other species. Many humans nowadays continuously collect activity data with their smart-watches or the phones. Thus, it would be interesting to evaluate the application of the DigiRhythm library in humans, offering an objective indicator of human welfare. A previous review highlights the potential of such rhythmicity computations in humans (https://journals.sagepub.com/doi/pdf/10.1177/0748730420940483).

At the date of writing this paper, the current version (v1.1) of the library does not offer the possibility to compute rhythmicity with other techniques such as cosinor analysis (\cite{refinetti2016}). Therefore, there is potential to expand the library's functionality in the future to include other techniques.

Finally, We invite other researchers to contribute to this library by submitting data, code, and techniques that can be shared with the scientific community. By collaborating and sharing these techniques, we can collectively advance our understanding of rhythmicity and its role in various biological systems.

>>>>

\section[Introduction]{Introduction} \label{sec:intro}

The real-time monitoring of animal behaviour is becoming increasingly important in agriculture. It can provide new opportunities to enhance the efficiency and sustainability of livestock production and is used to enhance animal welfare (Wemelsfelder et al., 2001). Animal behaviour can be altered by stress, pain or disease, but monitoring these changes in welfare is still a challenge, since some behavioural definitions are unclear and not specifically linked to a unique adverse event (Theurer et al., 2013; Levitis et al., 2009). Relevant information about animal welfare can be obtained by studying individual differences in the capacity for behavioural adaptation (Cooper et al., 2005). Indeed, during the animals life, significant changes can occur due to internal and external stimuli; these alterations arouse adaptive behavioural patterns that can be monitored by analysing the rhythmicity of behaviours (Berger et al., 2003).
Due to the increasing development of ICT (Information and Communication Technology) based on small, non-invasive sensors, large amounts of animal behaviour data are regularly recorded, and can be used to calculate parameters describing biological rhythms (Theurer et al., 2013; Scheibe and Gromann, 2006; Berger et al. 2003). Scheibe et al. (1999) developed a parameter able to quantify and investigate the rhythmicity of behavioural patterns, the Degree of Functional Coupling (DFC), which is based on telemetric measurements of activity and/or feeding. The DFC provides an estimate of synchronization between an individual and its environment. It is calculated to each 24 h period within rolling seven day periods and predicts values between $0\%$
and 100 $\%$ (Scheibe et al., 1999, Berger et al., 2003). An organism with a high DFC is highly synchronized with its environment, while lower DFC's indicate a lower synchronization. Breakdowns in the DFC response can be caused by stressors (e.g. transport, disease, etc.) that affect the time structure of behaviour (Berger et al., 2003; Scheibe et al., 1999). In fact, animals that live in stress-free conditions show activities all tuned to one zeitgeber (i. e. 24 hour periodicity), while their activity rhythm will deviate from the zeitgeber frequency if they are subjected to stressful situations (Scheibe et al., 1999; Berger et al., 2003).
Previous studies demonstrated that calves change their behaviour as illness develops (reduced feeding, less social behaviour and increased rest) as an adaptive response to cope with the disease (Owen-Ashley et al., 2006, Dantzer and Kelley, 2007; Borderas et al. 2008). A better understanding of changes in behaviour due to illness or stress of handling is valuable to improve animal welfare and productivity as well as handler safety (Grandin et al., 1989; Kilgore et al., 2005; Borderas et al., 2008; Robért et al., 2011). Bio-rhythmic analysis has the potential to provide the basis for the early detection of psychological and physiological disorders (Berger et al., 1999).

@@

The current package has been developed for farm animals, but the principles of rhythmicity apply for any organism, including humans. Therefore we highly recommend future research to evaluate the \pkg{digiRhythm} package in other species. Many humans nowadays continuously collect activity data with their smart-watches or the phones. Thus, it would be interesting to evaluate the application of the \pkg{digiRhythm} package in humans, offering an objective indicator of human welfare. A previous review highlights the potential of such rhythmicity computations in humans (https://journals.sagepub.com/doi/pdf/10.1177/0748730420940483).

At the time of writing this paper, the current version (v1.1) of the library does not provide the capability to compute rhythmicity using techniques such as cosinor analysis (\cite{refinetti2016}). However, there is potential for future expansion of the library's functionality to incorporate additional techniques. Researchers are encouraged to actively contribute by submitting new data, codes, or methods to the GitHub repository. These contributions will be reviewed by the package maintainers and, if approved, integrated into the \pkg{digiRhythm} package.
